The evolution of gene expression in the term placenta of viviparous mammals

### 2016-01-12 manuscript/manuscript.Rnw
fix monodelphis placenta ref

\begin{table}
  \tiny
<<top.50.opossum.expressed.genes,echo=FALSE,results="asis">>=
top.opossum.genes <- top.genes.table(species.name="monodelphis domestica")
print(xtable(top.opossum.genes),
      booktabs=TRUE,
      floating=FALSE,
      sanitize.text.function=genes.table.sanitize,include.rownames=FALSE)
setkey(top.opossum.genes,"Gene symbol")
@ 
\caption{\tableexplanation{\opossum}{tab:all_monodelphis_domestica}}
\label{tab:top50_opossum_placenta}
\end{table}

The top 50 expressed genes as measured by \ac{FPKM} in the term
placenta of the \opossum are shown in
\cref{tab:top50_opossum_placenta} with a list of all genes in
Supplemental \cref{tab:all_monodelphis_domestica}. The gene with the
second highest expression is superoxide dismutase 3, extracellular
(\Sexpr{gene.expression("monodelphis_domestica","SOD3",complete=TRUE)}),
which functions as an antioxidant. The gene with the second highest
expression is the hemoglobin subunit epsilon
(\Sexpr{gene.expression("monodelphis_domestica","ENSMODG00000011325",complete=TRUE)}),
which is expressed at the embryonic stage of development in
marsupials, as in other mammals
\cite{Holland.ea1997:Characterizationofembryonicglobin}. \textit{HBE}
subunits link up with protein products of two hemoglobin, zeta
(\textit{HBZ}) genes, which are also expressed to a high degree in the
opossum placenta
(\Sexpr{gene.expression("monodelphis_domestica","ENSMODG00000016635",complete=TRUE)}
and
\Sexpr{gene.expression("monodelphis_domestica","ENSMODG00000016631",complete=TRUE)}).
In addition, there is high expression of the gene hemoglobin, omega
(\textit{HBO},
\Sexpr{gene.expression("monodelphis_domestica","ENSMODG00000016637",complete=TRUE)}),
% this is LOC_1000016060 which I figured out by tblastn'ing AAK11486.1
% against Monodelphis domestica
a hemoglobin found only in marsupials and monotremes
\cite{Opazo.ea2008:Genomicevidenceforindependent}. The hemoglobin gene
\textit{HBA} is also expressed in the placenta
(\Sexpr{gene.expression("monodelphis_domestica","ENSMODG00000016633",complete=TRUE)}).
Two cystatin genes
(\Sexpr{gene.expression("monodelphis_domestica","ENSMODG00000018295",complete=TRUE)}
and
\Sexpr{gene.expression("monodelphis_domestica","ENSMODG00000018296",complete=TRUE)})
are expressed to a high degree, along with one \textit{CTSL} related
gene
(\Sexpr{gene.expression("monodelphis_domestica","ENSMODG00000002844",complete=TRUE)},
\cref{fig:cathespin_evolution_tree}), which are known to interact with
cystatins \cite{Turk.ea2001:Lysosomalcysteineproteasesfacts}. The
uncharacterized gene \textit{LOC10001195}
(\Sexpr{gene.expression("monodelphis_domestica","ENSMODG00000006547",complete=TRUE)})
also shows high expression in the opossum placenta. The gene \textit{WFDC2}, a
member of the whey acidic protein (WAP) family is also expressed to a
high degree in the opossum placenta
(\Sexpr{gene.expression("monodelphis_domestica","WFDC2",complete=TRUE)}).
Other genes with high expression include members of the annexin
complex
(\Sexpr{gene.expression("monodelphis_domestica","S100A6",complete=TRUE)},
\Sexpr{gene.expression("monodelphis_domestica","ANXA2",complete=TRUE)}).
In addition, there was little production of CYP19A1 and HSD3B1
(\Sexpr{gene.expression("monodelphis_domestica","CYP19A1",complete=TRUE)}
and
\Sexpr{gene.expression("monodelphis_domestica","ENSMODG00000005385",complete=TRUE)},
respectively).

@@

The
\Sexpr{core.placenta.transcriptome.long[,length(unique(human_name))]}
genes comprising the core placenta transcriptome are shown in
\cref{fig:core_placenta_transcriptome} ordered by their median
transcription level in \ac{FPKM}. We considered a gene to be core to
the placenta transcriptome if a 1:1 ortholog of that gene was
expressed with $\mathrm{FPKM} \ge 10$ in all species studied and was
not a human housekeeping gene according to
\citet{Eisenberg.ea2013:Humanhousekeepinggenesrevisited}.
\Cref{fig:core_placenta_transcriptome_shape} shows the number of genes
which would be included in the core placenta set if the \ac{FPKM}
cutoff was set differently.
%
Multiple components of the annexin complex are included in the core
transcriptome, including \textit{ANXA2}, \textit{ANXA1},
\textit{S100A11}, \textit{S100A10}. The modulation of  \DLA{It does more stuff; read
  the huthcinson2011 review and fill this in}
%
Another annexin, \textit{ANXA2} 
%
\textit{TIMP3}, which may be down-regulated in pre-eclampsia
\cite{Ma.ea2014:Down-regulationofTIMP3leads} (though another study
\cite{Xiang.ea2013:PromoterhypomethylationofTIMP3} claims
up-regulation)

@@

The opossum placenta shows expression of genes that either do not
exist in eutherians or are not found in the term eutherian placenta
(\cref{tab:top50_opossum_placenta}). The existence of embryonic forms
of hemoglobin in fetal tissues at term, as evidenced by the expression
of \textit{HBE} and \textit{HBZ} is a feature of marsupials that is
generally due to their very short gestation periods. Embryonic
hemoglobins in marsupials have a lower affinity for oxygen and allow
for a lower rate of oxygen uptake from maternal tissues
\cite{Holland.ea1997:Characterizationofembryonicglobin}. One novel
uncharacterized gene that shows high expression (LOC10001195) is not
found in eutherians; Ensembl gene trees reveal that homologs of this
gene exist in other amniotes such as birds and turtles, but are not
present in eutherian mammals. The homolog of this gene in chicken is
found to expressed in reproductive tract
\cite{Froman.ea2006:expressedsequencetaganalysis}, but its function is
unknown. \textit{WFDC2} is a member of the whey acidic protein (WAP)
family and involved in innate immunity. It is expressed in the mammary
gland of the tammar wallaby during pregnancy and early lactation
\cite{Sharp.ea2007:Molecularevolutionofmonotreme}, and here in
relative abundance in the opossum placenta.

>>>>

\begin{table}
  \tiny
<<top.50.opossum.expressed.genes,echo=FALSE,results="asis">>=
top.opossum.genes <- top.genes.table(species.name="monodelphis domestica")
print(xtable(top.opossum.genes),
      booktabs=TRUE,
      floating=FALSE,
      sanitize.text.function=genes.table.sanitize,include.rownames=FALSE)
setkey(top.opossum.genes,"Gene symbol")
@ 
\caption{\tableexplanation{\opossum}{tab:all_monodelphis_domestica_placenta}}
\label{tab:top50_opossum_placenta}
\end{table}

The top 50 expressed genes as measured by \ac{FPKM} in the term
placenta of the \opossum are shown in
\cref{tab:top50_opossum_placenta} with a list of all genes in
Supplemental \cref{tab:all_monodelphis_domestica_placenta}. The gene with the
second highest expression is superoxide dismutase 3, extracellular
(\Sexpr{gene.expression("monodelphis_domestica","SOD3",complete=TRUE)}),
which functions as an antioxidant. The gene with the second highest
expression is the hemoglobin subunit epsilon
(\Sexpr{gene.expression("monodelphis_domestica","ENSMODG00000011325",complete=TRUE)}),
which is expressed at the embryonic stage of development in
marsupials, as in other mammals
\cite{Holland.ea1997:Characterizationofembryonicglobin}. \textit{HBE}
subunits link up with protein products of two hemoglobin, zeta
(\textit{HBZ}) genes, which are also expressed to a high degree in the
opossum placenta
(\Sexpr{gene.expression("monodelphis_domestica","ENSMODG00000016635",complete=TRUE)}
and
\Sexpr{gene.expression("monodelphis_domestica","ENSMODG00000016631",complete=TRUE)}).
In addition, there is high expression of the gene hemoglobin, omega
(\textit{HBO},
\Sexpr{gene.expression("monodelphis_domestica","ENSMODG00000016637",complete=TRUE)}),
% this is LOC_1000016060 which I figured out by tblastn'ing AAK11486.1
% against Monodelphis domestica
a hemoglobin found only in marsupials and monotremes
\cite{Opazo.ea2008:Genomicevidenceforindependent}. The hemoglobin gene
\textit{HBA} is also expressed in the placenta
(\Sexpr{gene.expression("monodelphis_domestica","ENSMODG00000016633",complete=TRUE)}).
Two cystatin genes
(\Sexpr{gene.expression("monodelphis_domestica","ENSMODG00000018295",complete=TRUE)}
and
\Sexpr{gene.expression("monodelphis_domestica","ENSMODG00000018296",complete=TRUE)})
are expressed to a high degree, along with one \textit{CTSL} related
gene
(\Sexpr{gene.expression("monodelphis_domestica","ENSMODG00000002844",complete=TRUE)},
\cref{fig:cathespin_evolution_tree}), which are known to interact with
cystatins \cite{Turk.ea2001:Lysosomalcysteineproteasesfacts}. The
uncharacterized gene \textit{LOC10001195}
(\Sexpr{gene.expression("monodelphis_domestica","ENSMODG00000006547",complete=TRUE)})
also shows high expression in the opossum placenta. The gene \textit{WFDC2}, a
member of the whey acidic protein (WAP) family is also expressed to a
high degree in the opossum placenta
(\Sexpr{gene.expression("monodelphis_domestica","WFDC2",complete=TRUE)}).
Other genes with high expression include members of the annexin
complex
(\Sexpr{gene.expression("monodelphis_domestica","S100A6",complete=TRUE)},
\Sexpr{gene.expression("monodelphis_domestica","ANXA2",complete=TRUE)}).
In addition, there was little production of CYP19A1 and HSD3B1
(\Sexpr{gene.expression("monodelphis_domestica","CYP19A1",complete=TRUE)}
and
\Sexpr{gene.expression("monodelphis_domestica","ENSMODG00000005385",complete=TRUE)},
respectively).

@@

The
\Sexpr{core.placenta.transcriptome.long[,length(unique(human_name))]}
genes comprising the core placenta transcriptome are shown in
\cref{fig:core_placenta_transcriptome} ordered by their median
transcription level in \ac{FPKM}. We considered a gene to be core to
the placenta transcriptome if a 1:1 ortholog of that gene was
expressed with $\mathrm{FPKM} \ge 10$ in all species studied and was
not a human housekeeping gene according to
\citet{Eisenberg.ea2013:Humanhousekeepinggenesrevisited}.
\Cref{fig:core_placenta_transcriptome_shape} shows the number of genes
which would be included in the core placenta set if the \ac{FPKM}
cutoff was set differently.
%
Multiple components of the annexin complex are included in the core
transcriptome, including \textit{ANXA2}, \textit{ANXA1},
\textit{S100A11}, \textit{S100A10}. The modulation of  \DLA{It does more stuff; read
  the huthcinson2011 review and fill this in}
%
% Another annexin, \textit{ANXA2} 
%
\textit{TIMP3}, which may be down-regulated in pre-eclampsia
\cite{Ma.ea2014:Down-regulationofTIMP3leads} (though another study
\cite{Xiang.ea2013:PromoterhypomethylationofTIMP3} claims
up-regulation) does things

@@

The opossum placenta shows expression of genes that either do not
exist in eutherians or are not found in the term eutherian placenta
(\cref{tab:top50_opossum_placenta}). The existence of embryonic forms
of hemoglobin in fetal tissues at term, as evidenced by the expression
of \textit{HBE} and \textit{HBZ} is a feature of marsupials that is
generally due to their very short gestation periods. Embryonic
hemoglobins in marsupials have a lower affinity for oxygen and allow
for a lower rate of oxygen uptake from maternal tissues
\cite{Holland.ea1997:Characterizationofembryonicglobin}. One novel
uncharacterized gene that shows high expression (\textit{LOC10001195}) is not
found in eutherians; Ensembl gene trees reveal that homologs of this
gene exist in other amniotes such as birds and turtles, but are not
present in eutherian mammals. The homolog of this gene in chicken is
found to expressed in reproductive tract
\cite{Froman.ea2006:expressedsequencetaganalysis}, but its function is
unknown. \textit{WFDC2} is a member of the whey acidic protein (WAP)
family and involved in innate immunity. It is expressed in the mammary
gland of the tammar wallaby during pregnancy and early lactation
\cite{Sharp.ea2007:Molecularevolutionofmonotreme}, and here in
relative abundance in the opossum placenta.
